The noise-marginalized optimal statistic: A robust hybrid frequentist-Bayesian statistic for the stochastic gravitational-wave background in pulsar timing arrays

### 2018-05-14 manuscript/manuscript.tex
incorporated Steve's comments into manuscript

\title{The noise-marginalized optimal statistic: A robust hybrid frequentist-Bayesian statistic for a stochastic gravitational-wave background in pulsar timing arrays}
%\title{Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic}

@@

\author{Kristina P.\ Islo}
\affiliation{Center for Gravitation, Cosmology and Astrophysics, University of Wisconsin--Milwaukee, PO Box 413, Milwaukee WI, 53201, USA}

\author{Stephen R.\ Taylor}
\affiliation{Jet Propulsion Laboratory, California Institute of Technology, 4800 Oak Grove Drive, Pasadena, California 91106, USA}
\affiliation{Theoretical AstroPhysics Including Relativity (TAPIR), MC 350-17, California Institute of Technology, Pasadena, California 91125, USA}

@@

\begin{abstract}
Observations have revealed that nearly all galaxies 
contain supermassive black holes (SMBHs) at their centers. 
When galaxies merge, these SMBHs form SMBH binaries (SMBHBs), 
which emit low-frequency gravitational waves (GWs). 
The incoherent superposition of these sources produce a stochastic GW background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
The optimal statistic is a frequentist estimator of the amplitude of the GWB 
that specifically looks for the spatial correlations induced by the GWB in PTA data. 
In this paper, we introduce an improved method for computing the optimal statistic 
that marginalizes over the red noise in individual pulsars. 
We use simulations to demonstrate that this method 
more accurately determines the strength of the GWB, 
and we use the noise-marginalized optimal statistic 
to compare the significance of monopole, dipole, and Hellings-Downs spatial correlations 
and perform sky scrambles.
\end{abstract}

@@

Long-wavelength gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \cite{hd1983,fb1990}. 
The dominant astrophysical source in this frequency range is the isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the ultra-regular periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by searching for correlations in the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival. 
Current upper limits on the stochastic background from PTAs are approaching 
theoretical predictions for the GWB \citep{PPTA2013,EPTA2015,abb+17b}

PTAs primarily use Bayesian data analysis to compare the inferred probabilities 
of various models for the residuals, 
including one where they contain evidence of the GWB 
\citep{vlm+2009,lah+2013}. 
%Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated models. 
Bayesian inference is a powerful tool because 
it performs a non-linear fit and 
accounts for degeneracies between parameters. 
However, running a full Bayesian analysis is computationally intensive, 
particularly when searching for evidence of Hellings-Downs spatial correlations --
the ``smoking gun'' of the GWB.

@@

However, the optimal statistic gives biased results 
when the pulsars have significant red noise 
due to the strong covariance between the individual red noise parameters 
and the GWB amplitude. 
Many individual pulsars show evidence for red noise \citep{lam+2017,abb+17}, 
and uncertainty in the position of the Solar System barycenter leads to 
a common red process in all pulsars \citep{abb+17b}. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: 
marginalizing over the individual pulsar red noise parameters 
using the posterior distributions from a full Bayesian analysis of all the pulsars. 
This hybrid approach produces a more precise estimate of the GWB amplitude \Agw\ 
and its uncertainty, while requiring only a few minutes more than the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic for any choice of spatial correlations 
simply by changing the overlap-reduction function (ORF). 
This technique is used to perform the frequentist searches 
for common red signals with monopole, dipole, and Hellings-Downs spatial correlations 
in the NANOGrav 11-year data set \citep{abb+17b}.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic. 
We use simulations to compare the noise-marginalized optimal statistic 
to the optimal statistic computed with fixed noise. 
In Sec.~\ref{sec:spatial} we determine how well 
the noise-marginalized optimal statistic can 
differentiate between monopole, dipole, and Hellings-Downs spatial correlations. 
In Sec.~\ref{sec:skyscrambles} we use the noise-marginalized optimal statistic 
to perform sky scrambles, 
which assess the significance of Hellings-Downs spatial correlations 
by scrambling the pulsars' sky positions \citep{cs2016, tlb+2017}. 
We summarize our results in Sec.~\ref{sec:conclusion} 
and discuss future applications of the noise-marginalized optimal statistic.

@@

The optimal statistic is a frequentist estimator 
for the amplitude of an isotropic stochastic GWB, 
and can be derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime (see~\citet{abc+2009}). 
It is constructed from the timing residuals $\res$, 
which can be written as 
\begin{equation}
	\res = M {\bf{\epsilon}} + F {\bf{a}} + U {\bf{j}} + {\bf{n}} \,.
\end{equation}
The design matrix $M$ and timing model parameter uncertainties $\epsilon$ 
describe the contributions to the residuals from uncertainties in the timing model. 
The term $U{\bf{j}}$ describes noise that is correlated for observations made at the same time 
at different frequencies and uncorrelated over different observing epochs, 
while ${\bf{n}}$ describes uncorrelated white noise.
The term $F{\bf{a}}$ describes red noise, 
including both red noise intrinsic to the pulsar 
and a common red noise signal common to all the pulsar (such as a GW signal). 
We model the red noise as a Fourier series,
\begin{equation}
	F {\bf{a}} = \sum_{j=1}^N \left[ a_j \sin\left( \frac{2\pi j t}{T} \right) + b_j \cos \left( \frac{2\pi j t}{T} \right) \right] \,,
\end{equation}
where $N$ is the number of Fourier modes used (typically $N=30$) 
and $T$ is the span of the observations.

@@

The optimal statistic is constructed from the autocovariance and cross-covariance matrices 
$C_a$ and $S_{ab}$, 
\begin{eqnarray}
	C_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	S_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,,
\end{eqnarray}
where $\res_a$ is a vector of the residuals of the $a$th pulsar in the PTA. 
For the GWB with power spectrum $P_\mathrm{gw}(f)$ 
and overlap reduction function (ORF) $\Gamma_{ab}$, 
the cross-covariance matrices are
\begin{equation}
	S_{ab} = F_a \, \phi_{ab}^\mathrm{gw} \, F_b^T \,,
\end{equation}
where
\begin{equation}
	\phi_{ab}^\mathrm{gw} = \Gamma_{ab} \, P_\mathrm{gw}(f) \,,
	\label{eq:phi_gw}
\end{equation}
and the ORF is the Hellings-Downs curve \citep{hd1983}, 
\begin{eqnarray}
	\Gamma_{ab} &=& \frac{1}{2} \left[ 1 - \frac{1}{2} \left( \frac{1-\cos\theta_{ab}}{2} \right) \right. \nonumber \\
		&& \left. + 3 \left( \frac{1-\cos\theta_{ab}}{2} \right) \ln \left( \frac{1-\cos\theta_{ab}}{2} \right) \right]  \,,
\end{eqnarray}
where $\theta_{ab}$ is the angle between the pulsars. 
We model the power-spectrum of the GWB as a power law:
\begin{equation}
	P_\mathrm{gw}(f) = \Agw \left( \frac{f}{f_\mathrm{yr}} \right)^{-\gamma} \,,
\end{equation}
where $\gamma = 13/3$ assuming SMBHBs evolve solely due to GW emission and $f_\mathrm{yr} \equiv 1/(1\,\mathrm{yr})$. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{\sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) } \,,
\end{equation}
where $\tilde{S}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{S}_{ab} = S_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
If $\Agw=0$, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{-1/2} \,.
\end{equation}
For a measured value of $\hat{A}^2$, 
the significance of $\hat{A}^2 \neq 0$ is given by the signal-to-noise ratio (SNR) 
\begin{equation}
	\rho = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{ \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{1/2}} \,.
\end{equation}

When constructing the residuals $\res_a$, 
we typically fix the red noise parameters to the values that 
maximize the likelihood. 
However, this leads to a bias in the optimal statistic 
because the individual red noise and common red noise parameters 
are highly covariant. 
In this section, we compare three techniques for computing the optimal statistic. 
First, we fix the individual pulsars' red noise parameters to the 
maximum-likelihood values from individual Bayesian pulsar noise analyses. 
Second, we fix the pulsars' red noise parameters to the 
values that jointly maximize the likelihood for a Bayesian analysis 
of all of the pulsars in our PTA 
that searches over the pulsars' red noise parameters and 
a common red process. 
For the noise-marginalized method, we draw values of the pulsars' 
red noise parameters from the posteriors generated by the common Bayesian analysis.

We use these methods to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets 
consisting of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P_a(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details).
We use the PTA data analysis package 
PAL2 %\footnote{\href{https://github.com/jellis18/PAL2}{https://github.com/jellis18/PAL2}} 
\citep{evh17a}
to perform the noise analyses and compute the optimal statistic.

@@

The fixed-noise and noise-marginalized analyses give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing data sets 
containing smaller injected values of $\Agw$. 
In Fig.~\ref{fig:os_compare} we show a P--P plot 
of the cumulative fraction of simulations for which the injected $\Agw^2$ lies within 
a given confidence interval of the measured $\hat{A}^2$, 
assuming a Gaussian distribution with mean equal to the measured $\hat{A}^2$ 
and variance $\sigma^2_{\hat{A}^2}$, 
determined from the distribution of $\hat{A}^2$ for our simulations. 
If $\hat{A}^2$ has a Gaussian distribution centered around $\Agw^2$, 
the curves should lie along a straight line with slope equal to unity (the dotted, diagonal lines in Fig.~\ref{fig:os_compare}).

@@

Figure~\ref{fig:os_ORF} shows the noise-marginalized 
mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 200 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
Using quadrupole spatial correlations gives $\hat{A}^2 = (2.4 \pm 1.2) \times 10^{-29}$, 
while using monopole spatial correlations gives $\hat{A}^2 = (2.5 \pm 3.1) \times 10^{-30}$, 
and dipole spatial correlations gives $\hat{A}^2 = (5.3 \pm 4.2) \times 10^{-30}$. 
We find a noise-marginalized mean SNR above 1.0 in 97\% of our simulated data sets 
using the quadrupole ORF, and in 50\% and 68\% of our simulated data sets 
using the monopole and dipole ORFs, respectively. 
The mean SNR using the quadrupole ORF, averaged over realizations of the GWB, is 4.1, 
and we find an SNR greater than this using the monopole and dipole ORFs in just 
3\% and 3.5\% of our simulations, respectively.
\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_spatial_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 200 simulated data sets 
			containing an injected GWB with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), 
			and quadrupole (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

@@

The significance of spatial correlations can also be tested with ``sky scrambles,'' 
where the ORF is altered in order to simulate changing the pulsars' positions \citep{cs2016}. 
\citet{tlb+2017} showed how sky scrambles affect the Bayes' factor for simulated data sets. 
Here we perform a similar analysis using frequentist methods.

We generated 725 scrambled ORFs 
using a Monte Carlo algorithm. 
We required the scrambled ORFs to have $\bar{M} < 0.2$ 
with respect to the true ORF and each other. 
This threshold was chosen to be comparable to the match statistics 
between the Hellings-Downs ORF with monopole and dipole ORFs 
given in Sec.~\ref{sec:spatial}.
For each simulation, 
we computed the noise-marginalized mean optimal statistic and mean SNR 
for each scrambled ORF, and compared to the values found using the true ORF.

Figure~\ref{fig:skyscrambles_dataset_sample} shows the results of a sky scramble analysis 
for a sample data set with $\Agw=5\times10^{-15}$. 
For this particular realization of the GWB, 
none of the 725 scrambled ORFs resulted in a mean SNR greater than 
the mean SNR using the true ORF ($p < 0.0014$).
In Fig.~\ref{fig:skyscrambles_pvalues}, we plot the distribution of $p$-values of the 725 sky scrambles. 
For a GWB with $\Agw = 5\times10^{-15}$, 96\% of the simulations have $p \leq 0.05$, 
and 75\% of the simulations have $p \leq 0.003$. 
For a GWB with $\Agw = 10^{-15}$, 60\% of the simulations have $p \leq 0.05$, 
and 33\% have $p \leq 0.003$.
\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_scrambled_dataset11.pdf}
	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
			with and without sky scrambles for a simulated data set 
			containing a GWB with $\Agw = 5\times10^{-15}$. 
			None of the 725 scrambled ORFs gave a mean SNR as large 
			as the mean SNR using the true ORF ($p < 0.0014$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}

@@

The primary strength of the optimal statistic is how quickly it can be computed. 
This is useful for analyses where the significance of many spatial correlations are compared, 
as with the sky scrambles. 
It also makes the optimal statistic a valuable tool for analyzing simulations 
where many realizations of the GWB are compared. 
The noise marginalization technique described in this paper is key 
to being able to accurately measure the GWB with the optimal statistic 
for real PTAs and realistic PTA simulations, 
for which red noise is significant.

\acknowledgments
We thank Joe Romano, Xavier Siemens, Jolien Creighton, and Jeffrey Hazboun for useful discussions. 
This work was supported by NSF Physics Frontier Center Grant 1430284. 
SRT was partially supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA. 
JAE was partially supported by NASA through Einstein Fellowship Grant PF4-150120.

>>>>

\title{The noise-marginalized optimal statistic: A robust hybrid frequentist-Bayesian statistic for the stochastic gravitational-wave background in pulsar timing arrays}
%\title{Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic}

@@

\author{Kristina Islo}
\affiliation{Center for Gravitation, Cosmology and Astrophysics, University of Wisconsin--Milwaukee, PO Box 413, Milwaukee WI, 53201, USA}

\author{Stephen R.\ Taylor}
\affiliation{Theoretical AstroPhysics Including Relativity (TAPIR), MC 350-17, California Institute of Technology, Pasadena, California 91125, USA}
\affiliation{Jet Propulsion Laboratory, California Institute of Technology, 4800 Oak Grove Drive, Pasadena, California 91106, USA}

@@

\begin{abstract}
Observations have revealed that nearly all galaxies 
contain supermassive black holes (SMBHs) at their centers. 
When galaxies merge, these SMBHs form SMBH binaries (SMBHBs), 
which emit low-frequency gravitational waves (GWs). 
The incoherent superposition of these sources produce a stochastic GW background (GWB) 
that can be observed by pulsar timing arrays (PTAs). 
The optimal statistic is a frequentist estimator of the amplitude of the GWB 
that specifically looks for the spatial correlations between pulsars induced by the GWB. 
In this paper, we introduce an improved method for computing the optimal statistic 
that marginalizes over the red noise in individual pulsars. 
We use simulations to demonstrate that this method 
more accurately determines the strength of the GWB, 
and we use the noise-marginalized optimal statistic 
to compare the significance of monopole, dipole, and Hellings-Downs spatial correlations 
and perform sky scrambles.
\end{abstract}

@@

Long-wavelength gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7} \; \mathrm{Hz}$ can be observed with pulsar timing arrays (PTAs) 
composed of millisecond pulsars (MSPs) \cite{hd1983,fb1990}. 
The dominant astrophysical source in this frequency range is the isotropic stochastic 
gravitational wave background (GWB) 
made up of the incoherent superposition of GWs from inspiraling 
supermassive black hole binaries (SMBHBs) 
\citep{1995ApJ...446..543R, 2003ApJ...583..616J, 2003ApJ...590..691W}. 
By monitoring the periodic emission from these pulsars using radio telescopes, 
we can probe the dynamics of the spacetime through which the pulses travel. 
This is done by searching for correlations in the pulsar timing residuals, 
which measure the difference between the expected and observed pulse times of arrival. 
Current upper limits on the stochastic background from PTAs are approaching 
theoretical predictions for the GWB \citep{PPTA2013,EPTA2015,abb+17b}

PTAs primarily use Bayesian data analysis to compare the inferred probabilities 
of various models for the residuals, 
including one where they contain evidence of the GWB 
\citep{vlm+2009,lah+2013}. 
%Bayesian inference is a powerful technique because it allows us to compare arbitrarily complicated models. 
Bayesian inference is a powerful tool because 
it properly accounts for degeneracies between parameters 
and incorporate all sources of uncertainty into the analysis. 
However, running a full Bayesian analysis is computationally intensive, 
particularly when searching for evidence of Hellings-Downs (HD) spatial correlations --
the ``smoking gun'' of the GWB.

@@

However, when pulsars have significant red noise, 
the optimal statistic gives biased results 
due to the strong covariance 
between the individual red noise parameters and the GWB amplitude. 
Many individual pulsars show evidence for red noise \citep{lam+2017,abb+17}, 
and uncertainty in the position of the Solar System barycenter (SSB) leads to 
a common red process in all pulsars \citep{abb+17b}. 
%In this paper, we present a technique for improving the accuracy of the optimal statistic 
%by marginalizing the optimal statistic over the individual red noise parameters 
%using the posterior distributions from a full Bayesian analysis. 
%This hybrid approach produces a more accurate 
%estimate of \Agw\ and its uncertainty. 
Here we present a technique for improving the accuracy of the optimal statistic by including an additional step: 
marginalizing over the individual pulsar red noise parameters 
using the posterior distributions from a full Bayesian analysis of all the pulsars. 
This hybrid approach produces a more precise estimate of the GWB amplitude \Agw\ 
and its uncertainty, while requiring only a few minutes more than the more traditional method of computation. 
%This method is more computationally intensive than previous methods 
%for computing the optimal statistic, but still requires only minutes. 
%Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
%can be used to compute the optimal statistic with any choice of the ORF, 
%allowing additional spatial correlations to be tested very quickly.
Furthermore, the same Bayesian analysis drawn upon by the noise marginalization 
can be used to compute the optimal statistic for any choice of spatial correlations 
simply by changing the overlap-reduction function (ORF). 
For example, clock errors lead to a common red signal with monopole spatial correlations \citep{hcm+2012}, 
while uncertainty in the SSB produces dipole spatial correlations \citep{chm+2010}. 
This technique is used to perform the frequentist searches 
for common red signals with Hellings-Downs, 
monopole, and dipole spatial correlations 
in the NANOGrav 11-year data set \citep{abb+17b}.

This paper is organized as follows. In Sec.~\ref{sec:marg_os} 
we lay out the procedure for computing the noise-marginalized optimal statistic. 
We use simulations to compare the noise-marginalized optimal statistic 
to the optimal statistic computed with fixed noise. 
In Sec.~\ref{sec:spatial} we determine how well 
the noise-marginalized optimal statistic can 
differentiate between monopole, dipole, and Hellings-Downs spatial correlations. 
In Sec.~\ref{sec:skyscrambles} we use the noise-marginalized optimal statistic 
to perform sky scrambles, 
which assess the significance of Hellings-Downs spatial correlations 
by scrambling the pulsars' sky positions \citep{cs2016, tlb+2017}. 
We summarize our results in Sec.~\ref{sec:conclusion} 
as well as discuss future applications of the noise-marginalized optimal statistic.

@@

The optimal statistic is a frequentist estimator 
for the amplitude of an isotropic stochastic GWB, 
and can be derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime \citep{abc+2009,ccs+2015}. 
It is constructed from the timing residuals $\res$, 
which can be written as 
\begin{equation}
	\res = M {\bf{\epsilon}} + F {\bf{a}} + U {\bf{j}} + {\bf{n}} \,.
\end{equation}
The term $M\epsilon$ describes the contributions to the residuals 
from perturbations to the timing model. 
The term $U{\bf{j}}$ describes noise that is correlated for observations made at the same time 
at different frequencies and uncorrelated over different observing epochs (ECORR), 
while ${\bf{n}}$ describes uncorrelated white noise from TOA measurement uncertainties.
The term $F{\bf{a}}$ describes red noise, 
including both red noise intrinsic to the pulsar 
and a common red noise signal common to all pulsars (such as a GW signal). 
We model the red noise as a Fourier series,
\begin{equation}
	F {\bf{a}} = \sum_{j=1}^N \left[ a_j \sin\left( \frac{2\pi j t}{T} \right) + b_j \cos \left( \frac{2\pi j t}{T} \right) \right] \,,
\end{equation}
where $N$ is the number of Fourier modes used (typically $N=30$) 
and $T$ is the span of the observations.

@@

The optimal statistic is constructed from the autocovariance and cross-covariance matrices 
$C_a$ and $S_{ab}$, 
\begin{eqnarray}
	C_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	S_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,,
\end{eqnarray}
where $\res_a$ is a vector of the residuals of the $a$th pulsar in the PTA. 
For the GWB with power spectral density (PSD) $P_\mathrm{gw}(f)$ 
and overlap reduction function (ORF) $\Gamma_{ab}$, 
the cross-covariance matrices are
\begin{equation}
	S_{ab} = F_a \, \phi_{ab}^\mathrm{gw} \, F_b^T \,,
\end{equation}
where
\begin{equation}
	\phi_{ab}^\mathrm{gw} = \Gamma_{ab} \, P_\mathrm{gw}(f) \,,
	\label{eq:phi_gw}
\end{equation}
and the ORF is the Hellings-Downs curve \citep{hd1983}, 
\begin{eqnarray}
	\Gamma_{ab} &=& \frac{1}{2} \left[ 1 - \frac{1}{2} \left( \frac{1-\cos\theta_{ab}}{2} \right) \right. \nonumber \\
		&& \left. + 3 \left( \frac{1-\cos\theta_{ab}}{2} \right) \ln \left( \frac{1-\cos\theta_{ab}}{2} \right) \right]  \,,
\end{eqnarray}
where $\theta_{ab}$ is the angle between the pulsars. 
We model the PSD of the GWB as a power law:
\begin{equation}
	P_\mathrm{gw}(f) = \frac{\Agw^2}{12\pi^2} \left( \frac{f}{f_\mathrm{yr}} \right)^{-\gamma} \,,
\end{equation}
where $\gamma = 13/3$ assuming SMBHBs evolve solely due to GW emission and $f_\mathrm{yr} \equiv 1/(1\,\mathrm{yr})$. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{\sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) } \,,
\end{equation}
where $\tilde{S}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{S}_{ab} = S_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
If $\Agw=0$, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{-1/2} \,.
\end{equation}
For a measured value of $\hat{A}^2$, 
the significance of $\hat{A}^2 \neq 0$ is given by the signal-to-noise ratio (SNR) 
\begin{equation}
	\rho = \frac{\sum_{ab} \res_a^T {C_a^{-1} \tilde{S}_{ab} C_b^{-1} \res_b}}{ \left[ \sum_{ab} \Tr \left( C_a^{-1} \tilde{S}_{ab} C_b^{-1} \tilde{S}_{ba} \right) \right]^{1/2}} \,.
\end{equation}

When constructing the residuals $\res_a$, 
we typically fix the red noise parameters to the values that 
maximize the single-pulsar likelihood. 
However, this leads to a bias in the optimal statistic 
because the individual red noise and common red noise parameters 
are highly covariant, 
with the optimal statistic systematically lower than the true value of $\Agw^2$. 
In this section, we compare three techniques for computing the optimal statistic. 
First, we fix the individual pulsars' red noise parameters to the 
maximum-likelihood values from individual Bayesian pulsar noise analyses. 
Second, we fix the pulsars' red noise parameters to the 
values that jointly maximize the likelihood for a Bayesian analysis 
of all of the pulsars in our PTA 
that searches over the pulsars' red noise parameters and 
a common red process. 
For the noise-marginalized method, we draw values of the pulsars' 
red noise parameters from the posteriors generated by the common Bayesian analysis.

We use these methods to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets 
consisting of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, 
\begin{equation}
	P_a(f) = \frac{A_\mathrm{red}^2}{12\pi^2} \left( \frac{f}{f_\mathrm{yr}} \right)^{-\gamma} \,,
\end{equation}
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details).
We use the PTA data analysis package 
PAL2 %\footnote{\href{https://github.com/jellis18/PAL2}{https://github.com/jellis18/PAL2}} 
\citep{evh17a}
to perform the noise analyses and compute the optimal statistic.

@@

The fixed-noise and noise-marginalized analyses 
using the common noise results give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing data sets 
containing smaller injected values of $\Agw$. 
In Fig.~\ref{fig:os_compare} we show a P--P plot 
of the cumulative fraction of simulations for which the injected $\Agw^2$ lies within 
a given confidence interval of the measured $\hat{A}^2$. 
The confidence interval of $\hat{A}^2$ is determined 
assuming $\hat{A}^2$ follows a Gaussian distribution, 
with mean and variance $\sigma^2_{\hat{A}^2}$ 
taken from the distribution for $\hat{A}^2$ 
found from our 200 realizations of the GWB (i.e.,~the top panel of Fig.~\ref{fig:os_datasetstats}).
If $\hat{A}^2$ has a Gaussian distribution centered around $\Agw^2$, 
the curves should lie along a straight line with slope equal to unity (the dotted, diagonal lines in Fig.~\ref{fig:os_compare}).

@@

Figure~\ref{fig:os_ORF} shows the noise-marginalized 
mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 200 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
Using quadrupole spatial correlations gives $\hat{A}^2 = (2.4 \pm 1.2) \times 10^{-29}$, 
while using monopole spatial correlations gives $\hat{A}^2 = (2.5 \pm 3.1) \times 10^{-30}$, 
and dipole spatial correlations gives $\hat{A}^2 = (5.3 \pm 4.2) \times 10^{-30}$. 
We find a noise-marginalized mean SNR above 1.0 in 97\% of our simulated data sets 
using the quadrupole ORF, and in 50\% and 68\% of our simulated data sets 
using the monopole and dipole ORFs, respectively. 
The mean SNR using the quadrupole ORF, averaged over realizations of the GWB, is 4.1, 
and we find an SNR greater than this using the monopole and dipole ORFs in just 
3\% and 3.5\% of our simulations, respectively.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_spatial_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 200 simulated data sets 
			containing an injected GWB with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), 
			and quadrupole (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

@@

The significance of spatial correlations can also be tested with ``sky scrambles,'' 
where the ORF is altered in order to simulate changing the pulsars' positions \citep{cs2016,tlb+2017}. 
The scrambled ORFs are required to have small values of $\bar{M}$ 
with the true ORF and each other so that they form a nearly-orthogonal set. 
This ensures that the distribution of $\hat{A}^2$ computed using the scrambled ORFs 
forms the null hypothesis distribution. 
\citet{tlb+2017} showed how sky scrambles affect the Bayes' factor for simulated data sets. 
Here we perform a similar analysis using frequentist methods.

We generated 725 scrambled ORFs 
using a Monte Carlo algorithm. 
We required the scrambled ORFs to have $\bar{M} < 0.2$ 
with respect to the true ORF and each other. 
This threshold was chosen to be comparable to the match statistics 
between the Hellings-Downs ORF with monopole and dipole ORFs 
given in Sec.~\ref{sec:spatial}. 
We did not chose a smaller threshold because significantly more time 
would have been needed to generate 725 scrambled ORFs. 
For each simulation, we computed the 
noise-marginalized mean optimal statistic and mean SNR 
for each scrambled ORF, 
and compared them to the values found using the true ORF.

Figure~\ref{fig:skyscrambles_dataset_sample} shows the results of a sky scramble analysis 
for a sample data set with $\Agw=5\times10^{-15}$. 
For this particular realization of the GWB, 
none of the 725 scrambled ORFs resulted in a mean SNR greater than 
the mean SNR using the true ORF ($p < 0.0014$).
In Fig.~\ref{fig:skyscrambles_pvalues}, we plot the distribution of $p$-values of the 725 sky scrambles. 
For a GWB with $\Agw = 5\times10^{-15}$, 96\% of the simulations have $p \leq 0.05$, 
and 75\% of the simulations have $p \leq 0.003$. 
For a GWB with $\Agw = 10^{-15}$, 60\% of the simulations have $p \leq 0.05$, 
and 33\% have $p \leq 0.003$. 
This shows that for smaller values of \Agw, 
there is a greater chance that noise fluctuations will appear 
to have the spatial correlations of the GWB.
\begin{figure}[t]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_scrambled_dataset11.pdf}
	\caption{Comparison between the noise-marginalized mean optimal statistic and mean SNR 
			with and without sky scrambles for a simulated data set 
			containing a GWB with $\Agw = 5\times10^{-15}$. 
			None of the 725 scrambled ORFs gave a mean SNR as large 
			as the mean SNR using the true ORF ($p < 0.0014$).}
	\label{fig:skyscrambles_dataset_sample}
\end{figure}

@@

The primary strength of the optimal statistic is how quickly it can be computed. 
This is useful for analyses where the significance of many spatial correlations are compared, 
as with the sky scrambles. 
An upcoming paper will use the noise-marginalized optimal statistic 
to determine how well the spatial correlations corresponding to alternate GW polarizations 
can be measured. 
It also makes the optimal statistic a valuable tool for analyzing simulations 
where many realizations of the GWB are compared. 
The noise marginalization technique described in this paper is key 
to being able to accurately measure the GWB with the optimal statistic 
for real PTAs and realistic PTA simulations, 
for which red noise is significant.

\acknowledgments
We thank Joe Romano, Xavier Siemens, Jolien Creighton, and Jeffrey Hazboun for useful discussions. 
This work was supported by NSF Physics Frontier Center Grant 1430284. 
%SRT was partially supported by appointment to the NASA Postdoctoral Program 
%at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
%and the Universities Space Research Association through a contract with NASA. 
JAE was partially supported by NASA through Einstein Fellowship Grant PF4-150120.
